Analysing Suicide Event Patterns Show Trend amongst Crisis Calls

### 2024-01-23 outputs/paper/paper.qmd
Final changes to the file. Updated README. Entered the LLM usage. Final paper and pdf.

---
title: "In Depth Analysis of Event Patterns in Crisis Situations"
subtitle: "City of Toronto Data"
author: 
  - Vanshika Vanshika
thanks: "Code and data are available at: https://github.com/vanshikav2/Data_Analysis_1."
date: today
date-format: long
abstract: "Toronto Police Services' 10-year-long dataset on Persons in Crisis Calls for Service Attended was used to conduct an in-depth analysis of the major PIC call event: Suicide and its data characteristics. It is found that there is a direct relation/trend that can be observed between suicide PIC calls and several factors such as the specific time of the day, day of the week, and identifed hotspot geographical location. This is important as it's essential to analyze the patterns of suicide PIC calls to effectively deal with the problem at its root."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Toronto's Person In Crisis lines are an invaluable community asset that provides free, anonymous assistance to people experiencing a serious crisis. The frequency of crisis situations in today’s cities, particularly those involving populations at risk (PICs), highlights the need to better understand the dynamics behind such incidents. Over the years, the situation has drastically changed. As stated by Beverly Romeo-Beehler [@resource1], Before the COVID-19 pandemic hit in 2018 and 2019, the call centre handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines. This is quite a large number of calls being received. He further adds that calls for service fell to 1.7 million in 2020 and then slightly increased to 1.8 million in 2021. During the pandemic in 2020 and 2021, slightly more than 60% of emergency 9-1-1 calls were placed. The call centre received over 5,000 calls a day on average between 2018 and 2021, with nearly 3,000 of those calls being 9-1-1 calls.

The Toronto Police Service is in charge of the system, which is built to handle several kinds of crisis calls, each of which calls for a unique reaction. A variety of event types, including attempt suicide, person in crisis, elopee, jumper, overdose, and threatened suicide, are included in the dataset.The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014,2015. In this paper, all the neighborhood (Hood_158) data is used as an estimand to get a better understanding of all the hotspots of the PIC calls in Toronto neighborhood. In order to identify significant patterns and trends within the Persons in Crisis calls for help attended dataset, several examination of the interactions between type of events, geographical locations, month, day of the week & time of the day and event-specific aspect are used. The focus of this paper is to draw correlation between suicide related event and time of the day or day of the week.The analysis between suicide related events, their frequency and the geographical area of PIC call shows a definitive trend among these factors. The analysis aims to provide insightful information that will improve crisis response system and promote community well-being.

In this data section, various aspects of dataset comparisons are analyzed, and the process of data cleaning used to derive those datasets is discussed. The results section of the paper presents all the results derived from different comparisons and datasets, and their analysis, trends, and understandings are discussed in the discussion section. Lastly, the conclusion section concludes the paper and provides final insights. Cross-referencing sections and sub-sections is implemented for coherence and clarity.

@@

This dataset is collected and provided by the Toronto Police Service and the dataset was last updated January 11, 2024 as of January 21, 2024. It includes data from year 2014 to 2024. This dataset included event_id, the time,day,month, year of the call, the type of event that occured, if an apprehension after call was made, police division of event, whether an Occurrence was created or not. For the year 2014 and 2015, it also included the calls old and new name of neighbourhood, according to Hood_158 and Hood_140 system, which was left NSA after in order to protect the privacy of the parties.
Upon collection of the data, it was cleaned in several different ways to analyse the data fully. The columns of neighbourhood addresses were removed, including division and event id. Please see @tbl-Full_data for the sample of this dataset
```{r}
#| message: false
#| echo: false
#| label: tbl-Full_data
#| tbl-cap: Sample of Cleaned Person In Crisis for Calls Attended Data

# creates a table of the cleaned sample data set
head(cleaned_data, 5) |>
  kable(
    col.names = c("Id","Event Year", "Event Month", "Event DOW", "Event Hour", "Event Type", "Apprehension made"),
      booktabs = TRUE
  )
```
### Hood_158 and Type of Event 
Moreover, using the same data from Toronto Police Service, it was cleaned differently by deleting all the years after 2015 with no location, time,month,day of the event and adding hood_158 column of the data to show the geographical location of the calls in early years. A sample of this data set can be viewed by @tbl-Hood

@@

# creates a data set of the cleaned data set of the neighbourhoods
head(cleaned_data_hood, 5) |>
  kable(
    col.names = c("Id", "Event Type", "Hood_158 Number"),
      booktabs = TRUE
  )
```

@@

### Type of Events ###
From the years 2014 to 2023, Toronto has the highest number of PIC calls for the Person In crisis event which consists of all the different of type of crisis like elopee, jumper, and much more. The second highest category is suicide related which includes both suicide and suicide attempted and is a total of 91501 calls. The last category is Overdose.The total number of calls recorded in this dataset is 291991. The results can be seen @fig-events.

### Neighbourhood and Suicide PIC Calls ###
@fig-hood shows the relationship between neighborhood and suicide PIC calls. There are around 174 neighborhoods divided in Hood_158 system and the graph shows around 158 neighborhoods have reported a PIC call for suicide-related events. The most number of calls were received from Hood 170 which were about 1022 suicide related calls and second most calls from received from the area Hood_73 which were around 858 calls. The Hood_170 is Yonge-Bay
Corridor area and Hood_73 is Moss Park area in Old City of Toronto[@resource2].

@@

day_colors <- c("Sunday" = "red", "Monday" = "green", "Tuesday" = "blue", "Wednesday" = "purple",
                "Thursday" = "orange", "Friday" = "brown", "Saturday" = "pink")
# Create ggplot objects for days and time

@@

Upon thorough examination of the dataset, several significant findings emerge. Overall, suicide PIC calls show a very important and cruicial relationship with the other factors such as time of the day, day of the week, and geographical area. As highlighted in the Canadian Journal of Psychiatry[@resource3], suicide remains a noteworthy cause of death in Toronto, necessitating a nuanced understanding of the factors contributing to suicide and the development of targeted prevention strategies.

@@

# Convert the table to a data frame
#df <- as.data.frame.matrix(table_data)

# Reorder days of the week
#df$Day <- factor(row.names(df), levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

# Melt the data for ggplot
#melted_df <- melt(df, id.vars = "Day")

# Create a heatmap
#ggplot(melted_df, aes(x = variable, y = Day, fill = value)) +
 # geom_tile() +
#  scale_fill_gradient(low = "white", high = "darkred") +
 # labs(title = "Suicide Calls by Day of the Week and Hour of the Day",
#       x = "Hour of the Day",
 #      y = "Day of the Week") +
 # theme_minimal()

@@

However, it is crucial to acknowledge and discuss the inherent biases and limitations within the dataset. The presence of numerous "NAS" (Not A Statement) assigned rows in the raw data introduces the possibility of errors in the graphs and analyses, potentially diverting the accuracy of the findings. Additionally, the dataset lacks explicit information regarding the reasons for caller apprehension, hindering a direct correlation between apprehension and the nature of events.

>>>>

---
title: "Analysing Suicide Event Patterns Show Trend amongst Crisis Calls"
subtitle: "City of Toronto Data"
author: 
  - Vanshika Vanshika
thanks: "Code and data are available at: https://github.com/vanshikav2/Data_Analysis_1."
date: today
date-format: long
abstract: "Toronto Police Services' 10-year-long dataset on Persons in Crisis Calls for Service Attended was used to conduct an in-depth analysis of the major PIC call event: Suicide and its data characteristics. It is found that there is a direct relation/trend that can be observed between suicide PIC calls and several factors such as the specific time of the day, day of the week, and identifed hotspot geographical location. This is important as it's essential to analyze the patterns of suicide PIC calls to effectively deal with the problem at its root."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

Toronto's Person In Crisis lines are an invaluable community asset that provides free, anonymous assistance to people experiencing a serious crisis. The frequency of crises in today’s cities, particularly those involving populations at risk (PICs), highlights the need to better understand the dynamics behind such incidents. Over the years, the situation has drastically changed. As stated by Beverly Romeo-Beehler [@resource1], Before the COVID-19 pandemic hit in 2018 and 2019, the call center handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines. This is quite a large number of calls being received. He further adds that calls for service fell to 1.7 million in 2020 and then slightly increased to 1.8 million in 2021.

The Toronto Police Service is in charge of the system, which is built to handle several kinds of crisis calls, each of which calls for a unique reaction. A variety of event types, including attempt suicide, the person in crisis, elopee, jumper, overdose, and threatened suicide, are included in the data set. The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014,2015. In this paper, all the neighborhood (Hood_158) data is used as an estimand to get a better understanding of all the hotspots of the PIC calls in the Toronto neighborhood. To identify significant patterns and trends within the Persons in Crisis calls for help attended data set, several examinations of the interactions between type of events, geographical locations, month, day of the week & time of the day, and event-specific aspects are used. The focus of this paper is to draw the correlation between suicide-related events and the time of the day or day of the week. The analysis of suicide-related events, their frequency, and the geographical area of PIC calls shows a definitive trend among these factors. The analysis aims to provide insightful information that will improve the crisis response system and promote community well-being.

In this data section, various aspects of data set comparisons are analyzed, and the data-cleaning process used to derive those data sets is discussed. The results section of the paper presents all the results derived from different comparisons and data sets, and their analysis, trends, and understandings are discussed in the discussion section. Lastly, the conclusion section concludes the paper and provides final insights. Cross-referencing sections and sub-sections are implemented for coherence and clarity.

@@

This dataset is collected and provided by the Toronto Police Service and the dataset was last updated January 11, 2024, as of January 21, 2024. It includes data from the year 2014 to 2024. This dataset included event_id, the time, day, month, and year of the call, the type of event that occurred if an apprehension after the call was made, the police division of the event, and whether an Occurrence was created. For the years 2014 and 2015, it also included the calls for old and new names of neighborhood, according to the Hood_158 and Hood_140 system, which was left “NSA” after to protect the privacy of the parties.
Upon collection of the data, it was cleaned in several different ways to analyze the data fully. The columns of neighborhood addresses were removed, including division and event id. Please see @tbl-Full_data for the sample of this dataset
```{r}
#| message: false
#| echo: false
#| label: tbl-Full_data
#| tbl-cap: Sample of Cleaned Person In Crisis for Calls Attended Data

# creates a table of the cleaned sample data set
head(cleaned_data, 5) |>
  kable(
    col.names = c("Id","Event Year", "Event Month", "Event DOW", "Event Hour", "Event Type", "Apprehension made"),
      booktabs = TRUE
  )
```
### Hood_158 and Type of Event 
Moreover, using the same data from the Toronto Police Service, it was cleaned differently by deleting all the years after 2015 with no location,  time, month, or day of the event and adding the hood_158 column of the data to show the geographical location of the calls in early years. A sample of this data set can be viewed by @tbl-Hood.

@@

# creates a data set of the cleaned data set of the neighborhoods
head(cleaned_data_hood, 5) |>
  kable(
    col.names = c("Id", "Event Type", "Hood_158 Number"),
      booktabs = TRUE
  )
```

@@

### Type of Events ###
From the years 2014 to 2023, Toronto has the highest number of PIC calls for the Person In crisis events which consists of all the different types of crises like elopee, jumper, and much more. The second highest category is suicide related which includes both suicide and suicide attempted and is a total of 91501 calls. The last category is Overdose. The total number of calls recorded in this dataset is 291991. The results can be seen at @fig-events.

### Neighbourhood and Suicide PIC Calls ###
@fig-hood shows the relationship between neighborhood and suicide PIC calls. There are around 174 neighborhoods divided in the Hood_158 system and the graph shows around 158 neighborhoods have reported a PIC call for suicide-related events. The most number of calls were received from Hood 170 which were about 1022 suicide-related calls and the second most calls from received from the area Hood_73 which were around 858 calls. The Hood_170 is Yonge-Bay Corridor area and Hood_73 is Moss Park area in the Old City of Toronto[@resource2].

@@

day_colors <- c("Sunday" = "red", "Monday" = "green", "Tuesday" = "blue", "Wednesday" = "purple",
                "Thursday" = "orange", "Friday" = "brown", "Saturday" = "pink")

# Create ggplot objects for days and time
# code inspiration from: https://github.com/alyssaschleifer/ttc-subway-delay-analysis/blob/main/outputs/paper/Paper.Rmd

@@

Upon thorough examination of the data set, several significant findings emerge. Overall, suicide PIC calls show a very important and crucial relationship with the other factors such as time of the day, day of the week, and geographical area. As highlighted in the Canadian Journal of Psychiatry[@resource3], suicide remains a noteworthy cause of death in Toronto, necessitating a nuanced understanding of the factors contributing to suicide and the development of targeted prevention strategies.

@@

@@

However, it is crucial to acknowledge and discuss the inherent biases and limitations within the data set. The presence of numerous "NAS" (Not A Statement) assigned rows in the raw data introduces the possibility of errors in the graphs and analyses, potentially diverting the accuracy of the findings. Additionally, the data set lacks explicit information regarding the reasons for caller apprehension, hindering a direct correlation between apprehension and the nature of events.
